---
title: "Warm-up mini-Report Mosquitoes"
author: "FILL IN YOUR NAME"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

# BACKGROUND

Fill in some text here that provides background info on the WNV system, the blood meal DNA extractions, PCR, sequencing, etc.

```{r viremia, eval = FALSE}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# match colors to your species plot
cols_dur <- cols

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols_dur, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```
NOTE: Examples of data you can plot for the background info at https://github.com/saarman/BIOL3070/blob/main/Wk1_plot_viremia.md


# MAJOR QUESTION

Fill in here, the question we want to answer...

# MOTHODS

Fill in here, including overview of proceedure and methods used for this project.

## Fill in name of first analysis
```{r analysis-1}
# put code here
```

## Fill in name of second analysis/plot
```{r plot}
# put code for plotting here
```

# DISCUSSION

## Interpretation of analysis 1

## Interpretation of analysis/plot

# CONCLUSIONS